---
title: "Discussion Problems for Monte Carlo"
output: html_document
---

This is scrap content...  go below with caution. 

Let $X$ be the average of 1000 random variables "independent and identically distributed" (iid) random variables.  Using different distributions for the original 1000, make a histogram of 1,000,000 $X$'s.  

For this one, we don't need to check_if_X_in_A.  We just need to generate a bunch of $X$'s.  We can still use the same code...



```{r echo= F}
library(tidyverse)
library(purrr)

# First, write a function to simulate X
simulate_X = function(){
  a_bunch_of_random_variables = rbernoulli(10000, .3)
  # a_bunch_of_random_variables = rbernoulli(10000, .6)
  # a_bunch_of_random_variables = now do, binomial, geometric, and poisson.
  X = mean(a_bunch_of_random_variables)
  return(X)
}


# Now, we are going to do it lots of times.  
# Let's arrange the simulations in a data.frame with three columns
r = 10000
monte_carlo = data.frame(replicate = 1:r, 
                     X = rep(NA,r)) 
for(i in 1:r){
  monte_carlo$X[i] = simulate_X()
}

hist(monte_carlo$X,breaks = 80, freq = F)

Z = (monte_carlo$X - mean(monte_carlo$X))/sd(monte_carlo$X)
hist(Z,breaks = 80, freq = F)
xseq = seq(-4,4,len=100) 
lines(xseq,  dnorm(xseq))
```


What is this code doing?  If it isn't clear, fiddle with it.

```{r}
# Bernoulli:
pseq = seq(.01,.99,len = 100)
GeometricMean = rep(NA, 100)
for(i in 1:100){
  GeometricMean[i] = rgeom(10000,pseq[i]) %>% mean
}

plot(pseq, GeometricMean)
plot((1/pseq)-1, 
     GeometricMean, 
     log = "xy")
abline(0,1)
```


