---
title:  "STAT340: Discussion 1: Intro/setup"
author: "Name"
date:   "`r format(Sys.time(), '%d %B, %Y')`" # autogenerate date as date of last knit
documentclass: article
classoption: letterpaper
output:
  html_document:
    highlight: tango
    fig_caption: false
---

```{r setup, include=FALSE}
# if sourced, set working directory to file location
# added tryCatch in case knitting runs into error
tryCatch({
  if(Sys.getenv('RSTUDIO')=='1'){
    setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
  }}, error = function(e){}
)

# install necessary packages
if(!require(pacman)) install.packages("pacman")
pacman::p_load(knitr,tidyverse)

knitr::opts_chunk$set(tidy=FALSE,strip.white=FALSE,fig.align="center",comment=" #")
```

---

[Link to source file](https://raw.githubusercontent.com/karlrohe/340-Spring21/master/docs/discussion/01/ds01.Rmd)

## Warmup

Break off into pairs and introduce each other. Questions:

1. Year?
1. Major?
1. One type of data interested in studying?
   - (examples: simple numeric, text, time series, geospatial, network, survival, audio/video, other data specific to a subject area, etc...)
1. Some other fun/cool fact about you?

## Setup R/Rstudio

Note: Chromebooks do not support R/Rstudio and thus are not sufficient for this class. Please talk to instructors/TAs if you are unable to find a suitable computer to use for this course.

- [Install R](https://cloud.r-project.org/)
   - If you already have R, please check it's the most recent version (4.0.3)

```{r}
# show R version string
sessionInfo()$R.version$version.string
```

- [Install Rstudio](https://rstudio.com/products/rstudio/download/#download)

## Setup course folder

It's important to keep your course files organized. We recommend you organize your files in a folder structure like this:

 - STAT340
   - lecture
   - discussion
     - 01
     - 02
     - ...
   - homework
     - 01
     - 02
     - ...
   - project
     - data
     - analysis
     - plots
     - ...
   - ...

Take a minute now to make the above. You can place `STAT340` anywhere you like (e.g. Documents, Desktop, Dropbox, etc). Also move this file (`ds01.Rmd`) into `STAT340/discussion/01` . After knitting, the next line should show your `STAT340` directory structure correctly.

```{r, comment=""}
fs::dir_tree("../../../STAT340")
```

## Rstudio intro/refresher

 - Panels (source, console, environment/history, plots/help/viewer)
 - File operations (new, open, save, source, knit, etc.)
 - Working directory
 - Common shortcuts (run line, comment, insert chunk, etc.)

## Final check

Make sure your R and Rstudio are both installed, up to date, and ready to rock. If you're having issues, ask now!

## Optional: backup/version control

It is **strongly recommended** you use some kind of backup or version control system (GitHub/Lab, Dropbox, Onedrive, etc.) so you don't lose your course files if something unexpected happens. If you need help with this, please ask your instructors/TA.