
---
title: "Grading Schedule"
author: "Bi Cheng Wu"
date: "`r format(Sys.time(), '%d %B, %Y')`"
documentclass: article
classoption: letterpaper
output:
  html_document:
  highlight: tango
---

```{r setup, include=FALSE}
if(!require(pacman)) install.packages("pacman")
knitr::opts_chunk$set(tidy=FALSE,strip.white=FALSE,fig.align="center",comment=" #")
```

## Requirements

We need to create a fair grading schedule that is proportional to the appointment percentage of each grader (Karl doesn't need to grade since he's already spending a lot of time writing and teaching lectures):

|              |   |
|--------------|---|
| Jiaxin Hu    | 1 |
| Yongjoon Kim | 2 |
| Yanbo Shen   | 2 |
| Bi Cheng Wu  | 2 |
| Yongfeng Wu  | 2 |
| Bowen Zhang  | 2 |

### Creating assignments vector

Assignments schedule:

discussion 2, homework 1,    
discussion 3, homework 2,    
...    
discussion 12, homework 11

(there may be fewer than this, but this is the maximum number that Karl would assign)

```{r}
# construct vector of assignments
assignments = paste(rep(c("discussion","  homework"),times=11),
                    rep(2:12,each=2)-rep(0:1,times=11))
cat(assignments,sep="\n")
length(assignments)
```

### Create vector of graders

```{r}
graders = c(
  "Jiaxin Hu",
  rep("Yongjoon Kim",2),
  rep("Yanbo Shen",2),
  rep("Bi Cheng Wu",2),
  rep("Yongfeng Wu",2),
  rep("Bowen Zhang",2)
)
graders
length(graders)
```

We try to find an ordering that will give people an $N$ week gap. Of course, we can find this manually, but it's more fun to write it this way, where the next one is **totally randomly sampled** from the remaining graders.

Turns out $N=4$ is pretty easy to find.

```{r}
done = F
N = 4
all_graders = reordered = NULL
`%notin%` = Negate(`%in%`)

sampleNextGrader = function(remaining_graders,running_order){
    last_graders = na.omit(rev(running_order))[1:N]
    different_graders = remaining_graders[ remaining_graders %notin% last_graders]
    
    if(length(different_graders) == 0) stop()
    
    next_grader_place = sample(1:length(different_graders),1)
    next_grader = different_graders[next_grader_place]
    where_in_remaining = min(which(remaining_graders==next_grader))
    new_remaining = remaining_graders[-where_in_remaining]

    next_place = min(which(is.na(running_order)))
    running_order[next_place] = next_grader

    return(list("remaining"=new_remaining,"order"=running_order))
}

while(!done){
  
  tryCatch({
    all_graders <<- sample(rep(graders,2))
    reordered <<- rep(NA,length(all_graders))
    while(NA %in% reordered){
      res = sampleNextGrader(all_graders,reordered)
      all_graders <<- res$remaining
      reordered <<- res$order
    }
    
    final = data.frame(assignments,graders=reordered)
    
    for(G in unique(reordered)){
      if(0!=unname(diff(table(sub(" [^ ]*$","",final[final[2]==G,][[1]]))))) stop()
    }
    
    done <<- T
    
  }, error = function(e){
    # retry
  })
}

```

## Results

Here's the resulting distribution of grading duties:

```{r}
print(final,quote=F,row.names=F)
print(final[order(final[[2]]),],quote=F,row.names=F)
```

